---
title: "Excess deaths in Poland in 2020/21. An exploratory analysis"
author:
  - name: "Tomasz Przechlewski"
    email:  t.plata-przechlewski@psw.kwidzyn.edu.pl
    affiliation: "Powiślańska Szkoła Wyższa (Kwidzyn/Poland)"
description: (c) Tomasz Przechlewski / CC-BY license
date: "May 2021"
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message=F, warning=F)
## 
library("ggplot2")
library("dplyr")
library("scales")
library("ggthemes")
library("ggpubr")
library("tidyr")
library("readr")

picWd <- 12
picHt <- 8
spanV <- 0.5
spanVV <- 0.25
spanVVV <- 0.1
windowLen <- 12
mainColor <- "deeppink"
loessColor <- "steelblue"
mainBreaks <- "1 month"
unit1 <- 1000000
farbe19 <- '#F8766D'
farbe20 <- '#00BFC4'
farbe21 <- '#C77CFF'
firstYrWeek <- 1

wojcodes <- c('PL21', 'PL22', 'PL41', 'PL42', 'PL43', 'PL51', 'PL52',
'PL61', 'PL62', 'PL63', 'PL71', 'PL72', 'PL81', 'PL82', 'PL84', 'PL91', 'PL92')

```

## Introduction

There were a significant number of excess deaths in Poland in 2020
and still more were registerd in 2021. The phenomena is analyzed using statistical
exploratory analysis (Tukey 1980) and charts in particular.
We do not pretend to provide answers, we try to find
interesting or strange patters and ask questions...


## The scale of the problem

Assessing change in the number of deaths  as a difference
between 2015--2019 average and the current value one can easily find
that approx 80,000 more people died in Poland in connection
with or due to COVID19 (which present an unprecedented increase by 20%).
At the same time 29095 people died in connection with COVID19 in 2020,
with 27069 deaths in September--December alone (17 weeks or weeks number 36--52).
In the first 14 weeks of 2021 more 29309 were registerd. This gives
1711 and 2039 deaths per week respectively (in 2021 on the average there were
20% more deaths than in september-december 2020. BTW there were
1254559 new cases in weeks 1--14 of 2021, and 1248165 new cases in weeks 36--52 of 2020
(89,5 vs 73,4/week or approx 20% increase)

## The data

Deaths statistics come from the Eurostat database (table demo_r_mweek3: Deaths by week,
sex, 5-year age group and NUTS 3 region). The last reported week is 14th (starting Monday 2021-04-05).
COVID19 deaths statistics come from the announcements of the Ministry of Health
published at https://www.gov.pl/web/koronawirus/. Polish HM do not provide any database (sic!)
with COVID related data and until January 2021,
the messages were not even archived.
We archived them on our own.
Our data differs from the official ones because
our database is a copy of the daily messages and the Ministry of Health data
is corrected in the so-called in the meantime, often without any explanation or backtracking.
However the difference of number of deaths, is insignificantly small (according to our
base number of deaths classified as COVID19 victims as of April 30, 2021 is
67575 and according to the MZ 67501 messages
https://www.gov.pl/web/koronawirus/wykaz-zarazen-koronawirusem-sars-cov-2; difference
74 deaths or 0.1%).

Population data comes from *Bank Danych Lokalnych* (provided Central Statistical Office or CSOPL;
https://bdl.stat.gov.pl/BDL/dane/podgrup/temat.)
GDP data comes from Bank Danych Makroekonomicznych (provided by CSOPL;
https://stat.gov.pl/obszary-tematyczne/rachunki-narodowe/rachunki-regionalne/wstepne-szacunki-produktu-krajowego-brutto-w-przekrój-regionow-w-2019-roku,8,3.html)

NOTE1: There are 16 provinces in Poland (according to *National Register of Territorial Divisions of the Country* or TERYT) but 17 regions (EU NUTS classification) because the biggest
Polish **województwo Mazowieckie** (TERYT) is divided into two macroregions in NUTS.
Deaths statistics are published by GUS/Eurostat according to NUTS but COVID according to TERYT. To get
the common denominator we combined PL91 and PL92 macroregions (NUTS) back to
województwo Mazowieckie.
More information on NUTS is here
https://stat.gov.pl/en/regional-statistics/classification-of-territorial-units/classification-of-territorial-units-for-statistics-nuts/the-nuts-classification-in-poland/

NOTE2: As one can expect the provinces differs with area, population size, GDP/per capita,
health services availability, etc.
The biggest is Mazowieckie for example and it has the  highest
GDP per capita. The short summary of Polish provinces (for foreigners) is here:

```{r, warning=FALSE, message=F, echo=F}
library("knitr")
library("dplyr")
p <- read.csv(file='ludnosc_2019.csv', sep=';', header=T)
l <- read.csv(file='lekarze2019.csv', sep=';', header=T)
o <- read.csv(file='pkb_pc_2018.csv', sep=';', header=T)

p <- p %>% left_join(l, by="geo") %>% left_join(o, by="geo") %>%
  select (geo, pop, lekarze_pracujący_wg_podstawowego_miejsca_pracy, pkb) %>% as.data.frame()


#carsNo2019 <- sum(o$y2019)
#carsNo2016 <- sum(o$y2016)
#print (carsNo2019)
#o <- o %>% mutate(y2019f = y2019/carsNo2019 *100, y2016f = y2016/carsNo2016 *100 )
kable(p,
      col.names = c('name', 'population', 'doctors/per 10ths', 'GDP (% of average)'))
##o
```

BTW: province in Polish is *prowincja* (due to both are from Latin) but actually
Polish administrative provice is called "województwo", from "wodzić" -- ie commanding (the armed
troops in this context).  This is an old term/custom from the 14th century,
where Poland was divided into provinces (every province
ruled by a "wojewoda" ie chief of that province). More can be found
at Wikipedia (cf [Administrative divisions of Poland](https://en.wikipedia.org/wiki/Administrative_divisions_of_Poland))

## Numer of deaths related to COVID by province

```{r, warning=F}
### Zgony COVID (tygodniowe) ##

zc <- read.csv("MZNw.csv", sep = ';',  header=T, na.string="NA" )
## dane dot województw
ww <- read.csv("ludnosc_2019.csv", sep = ';',  header=T, na.string="NA" )
##

zc.mean <- zc %>% 
   filter(woj != "Polska" & ((year == 2021 & dow < 15) | (year == 2020))) %>%
   group_by(woj) %>%
   summarise(cc = sum(newd))
   
aveCovid = sum(zc.mean$cc)
ww <- ww %>% filter (geo != 'Polska')

avePop = sum(ww$pop)
aveCP = aveCovid / avePop * 1000000

zc1m <- left_join( zc.mean, ww, by=c("woj"="geo")) %>%
   mutate(cc = cc / pop * 1000000, ccp = cc/aveCP * 100) %>%
   select(woj,cc,ccp,pop)

## bar chart:

covid_by_province <- zc1m %>%
   ggplot(aes(x = reorder(woj, ccp ))) +
   geom_bar(aes(y = cc), stat="identity", alpha=.25, fill=default_violet ) +
   xlab(label="") +
   ylab(label="") +
   ggtitle("Deaths per 1mln 2020--2021 by province", subtitle='last week reported 14/2021') +
   ##
   geom_hline(yintercept = aveCP, color='red1', alpha=.4, size=.4) +
   theme(axis.text = element_text(size = 7)) +
   #theme(plot.title = element_text(hjust = 0.5)) +
   coord_flip()
   
covid_by_province_p <- zc1m %>%
   ggplot(aes(x = reorder(woj, ccp ))) +
   geom_bar(aes(y = ccp), stat="identity", alpha=.25, fill=default_violet ) +
   xlab(label="") +
   ylab(label="") +
   ggtitle("Deaths per 1mln 2020--2021 by province (as % of overall average)", 
           subtitle='last week reported 14/2021') +
   geom_hline(yintercept = 100, color='red1', alpha=.4, size=.4) +
   theme(axis.text = element_text(size = 7)) +
   #theme(plot.title = element_text(hjust = 0.5)) +
   coord_flip()


p1 <- ggarrange(covid_by_province, covid_by_province_p,  ncol = 1, nrow=2 )
p1
##ggsave(plot=p1, filename="COVID_by_woj.png", height = 9)
```

## Weekly number of deaths in Poland (2015--2021)

```{r, warning=F}
## Województwa i Polska łącznie
z_PL_Woj <- read.csv("demo_r_mweek3_PL.csv.gz", sep = ';',  header=T, na.string="NA" )
n <- read.csv("nuts.csv", sep = ';',  header=T, na.string="NA" )

## Tylko PL
z_PL <- z_PL_Woj %>% filter(geo == "PL") %>% as.data.frame()

## Prosty wykres dynamiki
## Od 2015
z_PL_tt <- z_PL %>% filter (sex == 'O' & age == 'T' & year > 2014)  %>% as.data.frame()

pgr <-   ggplot(z_PL_tt, aes(x= as.factor(week), y=value, group=as.factor(year), color=as.factor(year))) +
   geom_line(size=.5) +
   geom_point(size=2, alpha=.3) +
   scale_y_continuous(breaks=seq(2000, 20000, by=2000)) +
   xlab(label="week") +
   ylab(label="") +
   labs(caption="", color = "Rok") +
   #coord_polar(start=3*pi/2) +
   ggtitle("Total deaths weekly (Poland/2014--2021)")

pgr_polar <-   ggplot(z_PL_tt, aes(x= as.factor(week), y=value, group=as.factor(year), color=as.factor(year))) +
   geom_line(size=.5) +
   geom_point(size=2, alpha=.3) +
   scale_y_continuous(breaks=seq(2000, 20000, by=2000)) +
   xlab(label="week") +
   ylab(label="") +
   labs(caption="", color = "Rok") +
   coord_polar(start=3*pi/2) +
   ggtitle("Total deaths weekly (Poland/2014--2021)")
pgr_polar   
##ggsave(plot=pgr_polar, filename="COVID_PL_polar.png", height = 9)
```

BTW we resort to the same trick (polar coordinates) as used by Nightingale (1958) in her famous
book on Crimean War deaths.


## Number of deaths in Poland by province

Before 2020 slow linear increase in (almost; łódzkie/lubelskie are exceptions probably
due to migration) all provinces. 
Sharp increase in 2020.

```{r}
## Tylko województwa
## Połącz 91 + 92 (tylko w ramce województwa)
z_Woj <- z_PL_Woj %>% filter(geo %in% wojcodes) %>%
   mutate(geo=recode(geo, PL91="PL92")) %>%
   group_by(year, week, sex, age, geo) %>% 
   summarise(value=sum(value)) %>% as.data.frame

## Dodaj nazwy
z_Woj <- left_join(z_Woj, n, by='geo')

## Tylko ogółem płeć/wiek
z_Woj_T <- z_Woj %>% filter (sex == 'O' & age == 'T' & geo %in% wojcodes ) %>% as.data.frame()

z_Woj_T_sum <- z_Woj_T %>% group_by(year, geo) %>%
   summarise( mdeaths=sum(value, na.rm=T), na.rm=TRUE) %>%
   mutate (date = sprintf ("%i-01-01", year))  %>% as.data.frame()

## %% względem pierwszego roku
z_Woj_T_sumr <- z_Woj_T_sum %>% group_by(geo) %>%
   mutate(rm = mdeaths/first(mdeaths) * 100) %>% as.data.frame()
z_Woj_T_sumr <- left_join(z_Woj_T_sumr, n, by="geo")


c25 <- c(
   "dodgerblue2", 
   "#E31A1C", # red
   "green4",
   "#6A3D9A", # purple
   "#FF7F00", # orange
   "black", 
   "gold1",
   "skyblue2", 
   #"#FB9A99", # lt pink
   "palegreen2",
   "#CAB2D6", # lt purple
   "#FDBF6F", # lt orange
   "gray70", 
   "khaki2",
   "maroon", 
   "orchid1", 
   "deeppink1", 
   ############
   #"blue1", 
   #"steelblue4",
   #"darkturquoise", 
   #"green1", 
   #"yellow4", 
   #"yellow3",
   #"darkorange4", 
   "brown"
)

p2.a <- z_Woj_T_sumr %>% filter ( year < 2020) %>% 
   ggplot(aes(x=as.Date(date), y=rm, color=name)) +
   geom_smooth(method = "lm", se=F, size=.8, alpha=.5) +
   geom_point(size=.8) +
   scale_color_manual(values=c25) +
   #geom_text(aes(label=sprintf("%i", week), y= cdeaths), size=2.5) +
   #facet_wrap( ~geo, scales = "free_y") +
   xlab(label="") +
   ylab(label="") +
   scale_x_date( labels = date_format("%Y"), breaks ="1 year") +
   theme(plot.subtitle=element_text(size=9), legend.position="right")+
   ggtitle("Deaths per 1mln", subtitle="2015=100%" )


p2.b <- z_Woj_T_sumr %>% filter ( year < 2021) %>% 
   ggplot(aes(x=as.Date(date), y=rm, color=name)) +
   geom_smooth(method = "loess", se=F, size=.4, span = spanVV) +
   ##geom_smooth(method = "lm", se=F, size=.8, alpha=.5) +
   geom_point(size=.8) +
   scale_color_manual(values=c25) +
   #geom_text(aes(label=sprintf("%i", week), y= cdeaths), size=2.5) +
   #facet_wrap( ~geo, scales = "free_y") +
   xlab(label="") +
   ylab(label="") +
   scale_x_date( labels = date_format("%Y"), breaks ="1 year") +
   theme(plot.subtitle=element_text(size=9), legend.position="right")+
   ggtitle("Deaths per 1mln", subtitle="2015=100%" )

p2.ab <- ggarrange(p2.a, p2.b, ncol = 1, nrow = 2)
##ggsave(plot=p2.ab, "zgony_PL_x1r.png", width=picWd, height = picHt)
### [W2]
p2.ab
```


## Excess deaths by province

As mentioned earlier we measure **excess deaths** (or mortality)
as the difference between 2015–2019 average and the number of deaths registered
in relevant weeks in 2020 and 2021. This simple measure is commonly used,
ie https://ec.europa.eu/eurostat/statistics-explained/index.php?title=Excess_mortality_-_statistics

```{r, warning=F}
## średnia z lat 2015--2019 vs lata 2020/21

zz_0 <- z_Woj_T %>% filter(year > 2019) %>% as.data.frame()
#zz_0 <- left_join(zz_0, n, by="geo") #??#

zz15_19 <- z_Woj_T %>% filter(year < 2020) %>% 
   group_by(week, geo) %>%
   summarise( mdeaths=mean(value, na.rm=T), na.rm=TRUE) %>%
   as.data.frame()

## zgony + zgony covidowe
zz_x <- left_join(zz_0, zc, by = c('year'='year', 'week'='dow', 'name' = 'woj'))

## cdeaths od 1.10.2020 ##
## 35 tydzień
start_2nd_wave <- as.numeric(format(as.Date("2020-08-31"), "%U"))

## zgony+zgony covidowe + średnia 2015-19
zz_xx <- left_join(zz15_19, zz_x, by = c('week'='week', 'geo' = 'geo')) %>%
   select(year, week, geo, value, newc, newd, mdeaths) %>%
   mutate (cdeaths = newd/(value - mdeaths ) * 100 ) %>% 
   filter (week < 53 & week > start_2nd_wave ) %>%
   as.data.frame()
# Dodaj nazwy
zz_xx <- left_join(zz_xx, n, by='geo')

## Dodaj info o ludności
zz_xx_p <- left_join(zz_xx, ww, by=c('name'='geo'))

## Łącznie wojewódzkie
zz_xx_t <- zz_xx %>% group_by(name) %>%
   summarise(s_value = sum(value), s_nd = sum(newd), s_md = sum(mdeaths))

zz_xx_diff <- zz_xx_t %>% mutate (exdeaths = s_value - s_md) %>% 
   filter (name != 'Polska') %>%
   left_join(ww, by=c('name'='geo')) %>%
   mutate (exdp1m = exdeaths/pop * 1000000)

exd_total_PL <- sum(zz_xx_diff$exdeaths)
pop_total_PL <- sum(zz_xx_diff$pop)

exd_ave_PL <- exd_total_PL / pop_total_PL * 1000000

## bar chart

p3 <- zz_xx_diff %>%
   ggplot(aes(x = reorder(name, exdp1m ))) +
   geom_bar(aes(y = exdp1m), stat="identity", alpha=.25, fill=default_violet ) +
   xlab(label="") +
   ylab(label="") +
   ggtitle("Excessive deaths per 1mln 2020--2021 by province", 
           subtitle='last week reported 14/2021') +
   geom_hline(yintercept = exd_ave_PL, color='red1', alpha=.4, size=.4) +
   theme(axis.text = element_text(size = 7)) +
   coord_flip()

##ggsave(plot=p3, file= "exd_by_province.png")
## [W3]
p3
```


## Number of deaths in Poland 2020/21 by age groups

We can observe the pattern by various age gropus in the 
following chart (note different Y-scales)


```{r, warning=F}
##
## Płeć łącznie
z_PL_sex_o <- z_PL %>% filter ( sex == 'O') %>% as.data.frame()

z0 <- z_PL_sex_o %>% filter ( year >= 2015  & year < 2020 ) %>% as.data.frame
z1 <- z_PL_sex_o %>% filter ( year == 2020 ) %>% as.data.frame
z2 <- z_PL_sex_o %>% filter ( year == 2021 ) %>% as.data.frame

## średnie w okresie 1 -- (n-1)
zz0 <- z0 %>% group_by(age,week) %>% 
   summarise( year = 't19', vv = mean(value, na.rm=TRUE)) %>% as.data.frame
zz1 <- z1 %>% group_by(age,week) %>% 
   summarise( year = 't20', vv = mean(value, na.rm=TRUE)) %>% as.data.frame
zz2 <- z2 %>% group_by(age,week) %>% 
   summarise( year = 't21', vv = mean(value, na.rm=TRUE)) %>% as.data.frame

### Połącz
zz1 <- bind_rows(zz0, zz1, zz2)

p4 <- ggplot(zz1, aes(x=week, y=vv, color=year)) +
   geom_smooth(method="loess", se=F, span=spanV, size=.4) +
   geom_point(size=.4, alpha=.5) +
   facet_wrap( ~age, scales = "free_y") +
   xlab(label="") +
   ylab(label="") +
   ##theme_nikw()+
   theme(plot.subtitle=element_text(size=9), legend.position="top")+
   scale_color_manual(name="Year: ", 
                      labels = c("2015--2019 average", "2020", "2021"), 
                      values = c("t19"=farbe19, "t20"=farbe20, "t21"=farbe21 )) +
   ggtitle("Deaths by age groups")

##ggsave(plot=p4, "PL_deaths_by_agegrps.png", width=picWd)
p4
```

## Overall number of deaths in Poland 2020/21

Weekly total deaths as **percent of 2015--2019 province average**. Green vertical line denotes
week #47 (22-28/11/2020)

```{r, warning=F}
###
## wiek razem

z0 <- z_Woj %>% filter ( year >= 2015  & year < 2020 ) %>% as.data.frame
z1 <- z_Woj %>% filter ( year == 2020 ) %>% as.data.frame
z2 <- z_Woj %>% filter ( year == 2021 ) %>% as.data.frame

## średnie w okresie 1 -- (n-1)
zz0 <- z0 %>% group_by(name,week) %>% summarise( year = 't19', vv = mean(value, na.rm=TRUE)) %>% as.data.frame
zz1 <- z1 %>% group_by(name,week) %>% summarise( year = 't20', 
                                                 vv = mean(value, na.rm=TRUE)) %>% as.data.frame
zz2 <- z2 %>% group_by(name,week) %>% summarise( year = 't21', 
                                                 vv = mean(value, na.rm=TRUE)) %>% as.data.frame
### Połącz
zz1 <- bind_rows(zz0, zz1, zz2)
### Zamień na procenty

zz1p <- zz1 %>% pivot_wider(names_from = year, values_from = vv) %>%
   mutate (t20p = t20/t19 * 100, t21p = t21 / t19 * 100) %>%
   select(name, week, t20p, t21p) %>%
   pivot_longer(cols=c(t20p, t21p), names_to="year", values_to="vv") %>%
   as.data.frame()

lastWeek <- last(zz1$week)
## nie działa dla dwóch lat
## firstWeek <- lastWeek - windowLen
firstWeek <- firstYrWeek
zz1 <- zz1 %>% filter ( week >= firstWeek  ) %>% as.data.frame

p5 <- ggplot(zz1p, aes(x=week, y=vv, group=year, color=year)) +
   #geom_smooth(method="loess", se=F, span=spanVV, size=.4) +
   geom_line(size=.4, alpha=.5) +
   geom_point(size=.8, alpha=.5) +
   #facet_wrap( ~name, scales = "free_y") +
   facet_wrap( ~name, scales = "fixed") +
   xlab(label="") +
   ylab(label="") +
   geom_hline(yintercept = 200, color="firebrick", alpha=.3, size=0.4) +
   geom_hline(yintercept = 150, color="green1", alpha=.3, size=0.4) +
    ## 47 tydzień
   geom_vline(xintercept = 47, colour="green") +
   scale_x_continuous(breaks=seq(1, 52, by=5)) +
   theme(plot.subtitle=element_text(size=9),
         axis.text = element_text(size = 7),
         legend.position="top")+
   scale_color_manual(name="Year: ", labels = c("2020", "2021"), 
                      values = c("t20p"=farbe20, "t21p"=farbe21 )) +
   ggtitle("Deaths by province")

#ggsave(plot=p5, "zgony_PL_by_woj_O.png", width=picWd, height = picHt)
p5
```

Augmented version of the above.
Weekly Covid-related deaths as **percent of province average for 2020W36--2021W14** 
vs weekly total deaths
as **percent of 2015--2019 province average**. Green vertical line denotes
week #47 (22-28/11/2020)

```{r, warning=F}

## 36 tydzień =  1.09 początek drugiej fali
zc.mean <- zc %>% select (year, dow, woj, newd) %>%
   filter ((year == 2021 & dow < 15) | (year == 2020 & dow > 35)) %>%
   group_by(woj) %>%
   summarise( meand=mean(newd, na.rm=T)) %>% as.data.frame()
## na 1mln
#wojpop <- read.csv("ludnosc_2019.csv", sep = ';',  header=T, na.string="NA" )

zc.mean <- left_join(zc.mean, ww, by=c("woj"="geo"))
zc.mean <- zc.mean %>% mutate(meand = meand/pop * 1000000)

zc.pp <- left_join(zc, zc.mean, by="woj") %>%
   filter ((year == 2021 & dow < 15) | (year == 2020 & dow > 35)) %>%
   select (woj, dow, year, newd, meand) %>%
   mutate (vv=newd/meand * 100, year=sprintf("c%i", year - 2000)) %>%
   #select(woj, dow, year, vv, newd, meand) %>%
   select(woj, dow, year, vv) %>%
   rename(name=woj, week=dow) %>% as.data.frame()

## dodaj do info o zgonach
zz1zz <- bind_rows(zz1p, zc.pp)  %>%  filter (name != 'Polska')

p6 <- ggplot(zz1zz, aes(x=week, y=vv, group=year, color=year)) +
   #geom_smooth(method="loess", se=F, span=spanV, size=.4) +
   geom_line(size=.4, alpha=.5) +
   geom_point(size=.8, alpha=.5) +
   facet_wrap( ~name, scales = "free_y") +
   #facet_wrap( ~name, scales = "fixed") +
   xlab(label="") +
   ylab(label="") +
   #geom_hline(yintercept = 200, color="firebrick", alpha=.3, size=0.4) +
   #geom_hline(yintercept = 150, color="green1", alpha=.3, size=0.4) +
   ##theme_nikw()+
   scale_y_log10()+
   scale_x_continuous(breaks=seq(1, 52, by=5)) +
   geom_vline(xintercept = 47, colour="green") +
   ##labs(caption=source) +
   theme(plot.subtitle=element_text(size=9),
         axis.text = element_text(size = 7),
         legend.position="top")+
   #scale_color_manual(name="Year: ", labels = c("2020", "2021", "c20", "c21"), 
   #                   values = c("t20p"=farbe20, "t21p"=farbe21, "c19"=farbe19 )) +
   ggtitle("Deaths in Poland by province (weekly)")

## [W6]
##ggsave(plot=p6, "zgony_PL_by_woj_Ozz.png", width=picWd, height = picHt) 
p6
```

Question#1: why peaks of COVID/general deaths curves are shifted horizontally?

Question#2: Podkarpackie and Małopolskie recorded highest relative excess deaths in fatal late November weeks (over 250% of 2015-2019 average) yet both rank
low in all-time-loses table
(Podkarpackie #6, Małopolskie #15). Małopolskie even ranks last in the
number of excess deaths table.
HC failure or measurement artefact?


```{r, warning=F}
p6c <- zz1zz %>% filter ((week > 35 & (year == 't20p' | year =='c20'))   ) %>%
   ggplot(aes(x=week, y=vv, group=year, color=year)) +
   geom_line(size=.4, alpha=.5) +
   geom_point(size=.8, alpha=.5) +
   facet_wrap( ~name, scales = "free_y") +
   #facet_wrap( ~name, scales = "fixed") +
   xlab(label="") +
   ylab(label="") +
   #geom_hline(yintercept = 200, color="firebrick", alpha=.3, size=0.4) +
   #geom_hline(yintercept = 150, color="green1", alpha=.3, size=0.4) +
   ##theme_nikw()+
   scale_y_log10()+
   scale_x_continuous(breaks=seq(1, 52, by=2)) +
   geom_vline(xintercept = 47, colour="green") +
   ##labs(caption=source) +
   theme(plot.subtitle=element_text(size=9), 
         legend.position="top")+
   #scale_color_manual(name="Year: ", labels = c("2020", "2021", "c20", "c21"), 
   #                   values = c("t20p"=farbe20, "t21p"=farbe21, "c19"=farbe19 )) +
   ggtitle("Deaths in Poland by province (weekly)")

##ggsave(plot=p6c, "zgony_PL_by_woj_Occ.png", width=picWd, height = picHt) 
p6c

############################################################################
## Jeszcze raz ############################################################
# 36 tydzień =  1.09 początek drugiej fali
# średnie/max/min dla woj w okresie
zc.mean <- zc %>% select (year, dow, woj, newd) %>%
   filter ((year == 2021 & dow < 15) | (year == 2020 & dow > 35)) %>%
   group_by(woj) %>%
   summarise( min_d=min(newd), max_d=max(newd)) %>% as.data.frame()
## na 1mln
#wojpop <- read.csv("ludnosc_2019.csv", sep = ';',  header=T, na.string="NA" )

zc.mean <- left_join(zc.mean, ww, by=c("woj"="geo"))
zc.mean <- zc.mean %>% mutate(min_d = min_d/pop * 1000000, max_d = max_d/pop * 1000000)

zc.pp <- left_join(zc, zc.mean, by="woj") %>%
   filter ((year == 2021 & dow < 15) | (year == 2020 & dow > 35)) %>%
   select (woj, dow, year, newd, min_d, max_d) %>%
   mutate (vv=(newd - min_d)/(max_d - min_d), year=sprintf("c%i", year - 2000)) %>%
   #select(woj, dow, year, vv, newd, meand) %>%
   select(woj, dow, year, vv) %>%
   rename(name=woj, week=dow) %>% as.data.frame()

## dodaj do info o zgonach
zz1p_aggr <- zz1p %>%
   group_by(name) %>%
   summarise (min_v = min(vv, na.rm = T), max_v= max(vv, na.rm = T)) %>% as.data.frame()
zz1p <- left_join(zz1p, zz1p_aggr, by='name')

zz1p <- zz1p %>% mutate(vv=(vv - min_v)/(max_v - min_v)) %>% as.data.frame()

##zc.pp <- left_join(zz1p, zc.mean, by=c("name"="woj")) %>%
##   select(name, week, year, vv, min_v, max_v)
      
zz1zz <- bind_rows(zz1p, zc.pp)  %>%  filter (name != 'Polska')

p6a <- zz1zz %>% filter (week > 40) %>%
   ggplot(aes(x=week, y=vv, group=year, color=year)) +
   geom_line(size=.4, alpha=.5) +
   geom_point(size=.8, alpha=.5) +
   #facet_wrap( ~name, scales = "free_y") +
   facet_wrap( ~name, scales = "free_y") +
   xlab(label="") +
   ylab(label="") +
   #geom_hline(yintercept = 200, color="firebrick", alpha=.3, size=0.4) +
   #geom_hline(yintercept = 150, color="green1", alpha=.3, size=0.4) +
   ##theme_nikw()+
   #scale_y_log10()+
   ##labs(caption=source) +
   scale_x_continuous(breaks=seq(1, 52, by=2)) +
   theme(plot.subtitle=element_text(size=9), legend.position="top")+
   #scale_color_manual(name="Year: ", labels = c("2020", "2021", "c20", "c21"), 
   #                   values = c("t20p"=farbe20, "t21p"=farbe21, "c19"=farbe19 )) +
   ggtitle("Deaths in Poland by province: excess deaths vs covid per 1mln (weekly)",
           subtitle='Normalized')

## [W6]
##ggsave(plot=p6a, "zgony_PL_by_woj_Ozz_40.png", width=picWd, height = picHt)
## 
p6a
```

## COVID/excess deaths ratio

**Covid related deaths as % of excess deaths** (values of this ratio are increasing significantly. 
Because better tests are applied? :-) )

```{r, warning=F}
zz_0 <- z_Woj_T %>% filter(year > 2019) %>% as.data.frame()
##zz_0 <- left_join(zz_0, n, by='geo') ##??##

zz_15 <- z_Woj_T  %>% filter(year < 2020) %>% 
   group_by(week, geo) %>%
   summarise( mdeaths=mean(value, na.rm=T), na.rm=TRUE) %>%
   as.data.frame()

## zgony i zgony covidowe 2020-2021
zz_x <- left_join(zz_0, zc, by = c('year'='year', 'week'='dow', 'name' = 'woj'))

## cdeaths od 1.10.2020 ##
zz_xx <- left_join(zz_15, zz_x, by = c('week'='week', 'geo' = 'geo')) %>%
   select(year, week, geo, value, newc, newd, mdeaths) %>%
   mutate (cdeaths = newd/(value - mdeaths ) * 100 ) %>% 
   ##mutate(date=as.Date(paste(year, week, 1, sep="-"), "%Y-%U-%u")) %>%
   mutate(date= as.Date(sprintf("%i-%i-1", year, week), "%Y-%U-%u") ) %>%
   filter (as.Date(date) > "2020-09-30" ) %>%
   as.data.frame()
zz_xx <- left_join(zz_xx, n, by='geo')

p7 <- zz_xx %>% filter (geo != "Polska") %>%
   mutate(cdeaths = replace(cdeaths, which( cdeaths > 100 | cdeaths < -100), NA)) %>%
   ggplot(aes(x=as.Date(date), y=cdeaths, color=name)) +
   geom_smooth(method="loess", size=.4, se = F, span=spanV) +
   geom_point(size=.8) +
   xlab(label="yy/ww") +
   ylab(label="") +
   scale_x_date( labels = date_format("%y/%W"), breaks ="2 weeks") +
   ##coord_cartesian(ylim = c(0, 100)) +
   theme(plot.subtitle=element_text(size=9), legend.position="right")+
   ggtitle("COVID-related deaths/all deaths ratios" )

###[W7]
##ggsave(plot=p7, "zgony_PL_by_province_1.png", width=picWd)
p7
```

**Question#3**: why rate of COVID-related deaths to total of excess deaths increases with no obvious
relation to the stages/levels of COVID epidemics (as depicted below)?
To be precise: the rate was well below 50% in 2nd wave, very high between 2nd--3rd waves
(2021W7) and during 3rd wave...


```{r, warning=F}
##########
## Zgony covidowe na 1mln
zc1 <- left_join(zc, ww, by=c('woj' = 'geo')) %>%
   mutate (newd1m = newd / pop * unit1) %>% as.data.frame()

p8 <- zc1 %>% filter (woj != "Polska" ) %>%
   filter (year == 2020 | (year == 2021 & dow < 15 )) %>%
   ggplot(aes(x=as.Date(date), y=newd1m)) +
   geom_line(size=.4, color= default_cyan, alpha=.5) +
   #geom_point(size=.8, color=default_red, alpha=.5) +
   #geom_smooth(method="loess", size=.8, se = F, span=spanV) +
   geom_point(color= default_cyan, size=.8) +
   ##geom_text(aes(label=sprintf("%i", week), y= cdeaths), size=2.5) +
   facet_wrap( ~woj, scales = "fixed") +
   xlab(label="") +
   ylab(label="") +
   geom_hline(yintercept = 50, color="firebrick", alpha=.25, size=0.4) +
   geom_hline(yintercept = 100, color="firebrick", alpha=.25, size=0.4) +
   geom_vline(xintercept = as.Date("2020-11-22", format="%Y-%m-%d"), colour="firebrick", alpha=.25, size=0.8) +
   geom_vline(xintercept = as.Date("2021-02-15", format="%Y-%m-%d"), colour="firebrick", alpha=.25, size=0.4) +
   scale_x_date( labels = date_format("%W"), breaks ="4 weeks") +
   ##coord_cartesian(ylim = c(0, 100)) +
   theme(plot.subtitle=element_text(size=9),
         axis.text = element_text(size = 6),
         legend.position="right")+
   ggtitle("Covid19 related deaths per 1mln" )

##ggsave(plot=p8, "zgony_PL_covid_O.png", width=picWd)
###[W8]
p8
```

## Excess Deaths vs GDP/access to healthcare 

Any correlation with GDP or health care availability? The measure of HC availibility is
*number of doctors per 10 ths population* 
(lekarze pracujący wg podstawowego miejsca pracy na 10 tys. ludności
doctors employed by the primary workplace per 10 thous. population; cf: https://bdl.stat.gov.pl/BDL/metadane/cechy/3173)

```{r, warning=F}
lekarze <-  read.csv("lekarze2019.csv", sep = ';',  header=T, na.string="NA" )
pkb <- read.csv("pkb_pc_2018.csv", sep = ';',  header=T, na.string="NA" )

zy_s_diff_ex <- zz_xx_diff %>% 
   left_join(lekarze, by=c("name"="geo")) %>%
   left_join(pkb, by=c("name"="geo")) %>%
   rename (lekarze = lekarze_pracujący_wg_podstawowego_miejsca_pracy) %>%
   mutate (lekarze1m = lekarze/pop * 1000000)


p_exd_vs_lekarze <- ggplot(zy_s_diff_ex, aes(x = lekarze1m, y = exdp1m)) +
   geom_point(size = 1) +
   geom_smooth(method = "lm") +
   geom_smooth(method = "loess", color='red')
p_exd_vs_lekarze 


###
lm <- lm(data=zy_s_diff_ex, exdp1m ~ lekarze1m );
summary(lm)

p_exd_vs_gdp <- ggplot(zy_s_diff_ex, aes(x = pkb, y = exdp1m)) +
   geom_point(size = 1) +
   geom_smooth(method = "lm") +
   geom_smooth(method = "loess", color='red')
p_exd_vs_gdp

lm2 <- lm(data=zy_s_diff_ex, exdp1m ~ pkb );
summary(lm2)

p_lekarze_vs_gdp <- ggplot(zy_s_diff_ex, aes(x = pkb, y = lekarze)) +
   geom_point(size = 1) +
   geom_smooth(method = "lm") +
   geom_smooth(method = "loess", color='red')
p_lekarze_vs_gdp

p9 <- ggarrange(p_exd_vs_lekarze, p_exd_vs_gdp, p_lekarze_vs_gdp, ncol=3, nrow=1)

###[W9]
##ggsave(plot=p_corr, "zgony_vs_lekarze_gdp.png", width=picWd)
p9
```

Doctors vs ED:  coefficient = 14.874 (probab 0,053 barely significant and R² = 24%) 
GDP vs ED: coefficient -2,39 (prob 0,23 and R² = 10%)

One additional doctor/10 ths or 100 additional doctors per 1 mln population saves 14 lives,
so at least statistically it is worth to invest in HC :-)



## Technical remarks

This document is an abridged version of a report 
prepared as R-markdown document (Xie 2021). 
The source file and the data are available at author's github repo
here: 

## Bibliography

Nightingale Florence,
*Mortality of the British Army, at Home and Abroad, 
and During the Russian War as Compared with the Civil Population in England*, Harrison&Sons, London 1858, (https://archive.org/details/mortalityofbriti00lond)

Tukey John W., We Need Both Exploratory and Confirmatory
*The American Statistician*
Vol. 34, No. 1 (Feb., 1980), pp. 23-25 (3 pages)

Xie Yihui, Allaire J.J,  Grolemund Garrett,
*R Markdown: The Definitive Guide*
https://bookdown.org/yihui/rmarkdown/
2021-04-09
